Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-10-02 Dail_Mad_ext.tex
discussion

\begin{comment}
  The model (1) recognizes
  the discrete nature of count data and the underlying true abundance
  state, (2) includes a realistic observation model with explicit
  detection probability parameters (3) has estimable parameters,
  including those describing the initial state at time $t=1$, and (4)
  allows for inference about both temporal and spatial variation in
  abundance.

\end{comment}

Our work has highlighted the limitations of classical state-space
models as applied to ecological time series data, and we have
demonstrated how open population $N$-mixture models overcome these
limitations. We extended this class of models in several important
ways that should make them more suitable to typical ecological
datasets that are characterized by sparse counts, zero-inflation, and
observation error due to imperfect detection.

We made four important developments of the open population
N-mixture model proposed by \citep{dail_madsen:2011} that make the
model more applicable to long-term data such as is collected
by the North American Breeding Bird Survey. First, we have
reconciled the objectives of traditional state-space models
with open population $N$-mixture models by illustrating how
classical models of population dynamics can be embedded within
the framework. Second, we have demonstrated methods of
accounting for zero-inflation in the time-series. Third, we
have illustrated how additional random effects such as
observer-specific detection probabilities can be
accommodated. Fourth, we presented a Bayesian analysis of the
model, which, makes it possible to incorporate prior information when
available, and in some cases, facilitates parameter estimation.

A unique aspect of the Dail and Madsen model as originally
devised was that it allowed for the estimation of demographic
parameters under strict distributional assumptions and with the
assumption of geographic closure. Clearly, estimating
demographic parameters from count data is an ambitious goal, and
the required assumption will not be valid in many cases. One
advantage of the original DM models over our extensions here is
that they model population dynamics more mechanistically, by
focusing on survival and recruitment. As we report here, these
models often provide unrealistic estimates of these vital rates
and are generally less supported in model selection. Adding
density dependence directly to one or both vital rates may solve
both of these problems, although this might require very good
data. Another strategy would be to use estimates of vital rates
from demographic studies as priors in a Bayesian analysis using
DM models. Finally, count data could be combined with other data
types such as capture-recapture data, which offers a means of
making inferences about demographic processes at broad spatial
scales. Models combining count data and demographic data are
often referred to as integrated population models (IPM; CITE),
and we believe that a DM model would provide an effective part
of an IPM.

In contrast, recent work has sought to use state-space
models for inference about population processes such as mortality and
recruitment. To do so, methods have been developed to combine multiple
sources of information, such as counts and mark-recapture data
\citep{besbeas_etal:2002, buckland_etal:2004,
  schaub_etal:2007}. Although these integrated
population models may be viewed as the gold standard in state-space
modeling, ecologists are not always so fortunate to have direct
information about vital rates, especially at large spatial scales.
Rather, count data are much more common, and
are produced by many of the largest monitoring programs in the world,
such as the North American Breeding Bird Survey
\citep[BBS;][]{robbins_etal:1986}.

>>>>

\begin{comment}
  The model (1) recognizes
  the discrete nature of count data and the underlying true abundance
  state, (2) includes a realistic observation model with explicit
  detection probability parameters (3) has estimable parameters,
  including those describing the initial state at time $t=1$, and (4)
  allows for inference about both temporal and spatial variation in
  abundance.
\end{comment}

Our work has highlighted the limitations of classical state-space
models as applied to ecological time series data, and we have
demonstrated how a class of open population $N$-mixture models
proposed by \citet{dail_madsen:2011} overcomes these
limitations. We extended this class of models in several important
ways to accomodate common features of ecological
datasets, including sparse counts, zero-inflation, demographic
stochasticity, and random variation in observation error.
%We made four important developments of the open population
%N-mixture model proposed by \citet{dail_madsen:2011} that make the
%model more applicable to long-term data such as is collected
%by the North American Breeding Bird Survey.
%First, we have
%reconciled the objectives of traditional state-space models
%with open population $N$-mixture models by illustrating how
%classical models of population dynamics can be embedded within
%the framework. Second, we have demonstrated methods of
%accounting for zero-inflation in the time-series. Third, we
%have illustrated how additional random effects such as
%observer-specific detection probabilities can be
%accommodated. Fourth, we presented a Bayesian analysis of the
%model, which, makes it possible to incorporate prior information when
%available, and in some cases, facilitates parameter estimation.

We also demonstrated how many of the objectives of classical
state-space models can be met using this. For example, one of the
primary aims of our paper was to demonstrate how classical
population growth models can be embedded in the DM model.
In some ways, such population growth models are more phenomenological
than mechanistic, and this runs counter to the motivation for the DM
model. However,
%A unique aspect of the Dail and Madsen model as originally
%devised was that it allowed for the estimation of demographic
%parameters under strict distributional assumptions and with the
%assumption of geographic closure. Clearly,
estimating
demographic parameters from count data is an ambitious goal, and
the required assumption will not be valid in many cases, especially
when immigration and emigration occur. This was supported by the
unrealistic estimates of vital rates reported here and by the lack of
support for the mechanistic models relative to the population growth
models.
%One
%advantage of the original DM models over our extensions here is
%that they model population dynamics more mechanistically, by
%focusing on survival and recruitment. As we report here, these
%models often provide unrealistic estimates of these vital rates
%and are generally less supported in model selection.
%Adding
%density dependence directly to one or both vital rates may solve
%both of these problems, although this might require very good
%data.
One strategy for adhereing to the explicit model would be to use
estimates of vital rates from demographic studies as priors in a
Bayesian analysis. Alternatively, count data could be combined with
other data types such as capture-recapture data, which offers a means of
making inferences about demographic processes at broad spatial
scales. Models that are fitted to count data and demographic data are
often referred to as integrated population models
\citep[][IPM]{besbeas_etal:2002, buckland_etal:2004,schaub_etal:2007}
and we believe that a DM model would provide an framework for an
IPM. However, we caution that when count data are affected by movement
and vital rates, IPMs would need to model

%In contrast, recent work has sought to use state-space
%models for inference about population processes such as mortality and
%recruitment. To do so, methods have been developed to combine multiple
%sources of information, such as counts and mark-recapture data
%\citep{besbeas_etal:2002, buckland_etal:2004,
%  schaub_etal:2007}.

Although these integrated
population models may be viewed as the gold standard in state-space
modeling, ecologists are not always so fortunate to have direct
information about vital rates, especially at large spatial scales.
Rather, count data are much more common, and
are produced by many of the largest monitoring programs in the world,
such as the North American Breeding Bird Survey
\citep[BBS;][]{robbins_etal:1986}.
